Conductive and convective heat transfer in inductive heating of subsea buried pipelines

### 2021-12-21 manuscript.tex
:pencil: Corrections till fig. 7

As the permeability decreases from $k =$ \SI{4.9E-10} to \SI{1.6E-11}{\meter\squared}, the heat distributes more uniformly around the cable, heating the pipe and the fluid; but still showing a preferential upward heat distribution reaching the ground surface typical in a convection-dominated behavior. The pipe temperature increases showing symmetric heat distribution around the cable as the permeability decreases from \SI{1E-12} to \SI{1E-15}{\meter\squared}. In these low permeability conditions, conduction dominates, and the pipeline is heated efficiently, barely losing any heat to the surrounding soil. At very low permeabilities (in silty and clayey soil with $k < \SI{1E-12}{\meter\squared}$ and $d_m < \SI{70}{\micro\meter}$), the narrow and sparsely interconnected pore spaces limit the fluid flow velocity. Hence, the time scale of convection is much larger than that of conduction, i.e., a smaller Rayleigh number indicating conduction-driven flows. Conductive heat transfer causes an approximately isotropic distribution of heat around the source.

@@

\begin{figure}
 \centering
 \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{figs/ra/955.png}
    \caption{$Ra$ = 955, $k$ = \SI{1.91E-9}{\meter\squared}}
 \end{subfigure}
 \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{figs/ra/269.png}
    \caption{$Ra$ = 269.5, $k$ = \SI{5.64E-10}{\meter\squared}}
 \end{subfigure}\\
 \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{figs/ra/112.png}
    \caption{$Ra$ = 112, $k$ = \SI{2.47E-10}{\meter\squared}}
 \end{subfigure}
 \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{figs/ra/21.png}
    \caption{$Ra$ = 21.95, $k$ = \SI{4.89E-11}{\meter\squared}}
 \end{subfigure}\\ 
 \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{figs/ra/7.png}
    \caption{$Ra$ = 7.29, $k$ = \SI{1.67E-11}{\meter\squared}}
 \end{subfigure}
 \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{figs/ra/2.png}
    \caption{$Ra$ = 2.225, $k$ = \SI{4.84E-12}{\meter\squared}}
 \end{subfigure}\\  
\end{figure}%
\begin{figure}[ht]\ContinuedFloat 
 \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{figs/ra/0p45.png}
    \caption{$Ra$ = \SI{4.58E-1}, $k$ = \SI{1.05E-12}{\meter\squared}}
 \end{subfigure}
 \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{figs/ra/0p049.png}
    \caption{$Ra$ = \SI{4.9E-2}, $k$ = \SI{1.11E-13}{\meter\squared}}
 \end{subfigure}\\    
 \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{figs/ra/0p0049.png}
    \caption{$Ra$ = \SI{4.4E-4}, $k$ = \SI{1.05E-14}{\meter\squared}}
 \end{subfigure}
 \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{figs/ra/0p00049.png}
    \caption{$Ra$ = \SI{4.4E-4}, $k$ = \SI{1.18E-15}{\meter\squared}}
 \end{subfigure}\\ 
 \caption{Temperature distribution in different soil permeability conditions. The cables are heated to an above ambient temperature of \SI{70}{\celsius} and the burial depth is \SI{0.5}{\meter}.}
 \label{fig:ra-evol}
\end{figure}

The heat transfer mechanism changes from conduction to convection as the permeability of soil increases. A critical Rayleigh number defines this transition point from conduction to convection.~\citeN{nield1968onset} defines this transition at a critical $Ra$ of 12 for a rectangular domain with a constant heat source with Free-Free boundary condition. At low permeabilities, where $Ra < 1$, the temperature field is symmetric, indicating a conductive behavior (see~\cref{fig:ra-evol}). The asymmetric convective behavior occurs for $Ra > 2.22$ ($k > \SI{4.84E-12}{\meter\squared}$). The observation of convective response for $Ra = 2.22$, below the critical $Ra$ of 12, indicates the transition occurs earlier than originally thought.

\Cref{fig:ra-vline} shows the temperature distribution with depth along the A-A' axis for different Rayleigh numbers. The fluid temperature in the pipe reaches a maximum of 73.5\% of the cable temperature for the lowest $Ra = $\SI{4.5E-4} ($k =$ \SI{1.18E-15}{\meter\squared}). The fluid temperature inside the pipe remains almost unchanged with increasing Rayleigh number until a $Ra = 0.46$ ($k >$ \SI{1E-12}{\meter\squared}), above which the fluid temperature drops drastically. For large Rayleigh numbers ($Ra > 100$), the fluid in the pipe barely reaches 5\% of the cable temperature. Below the pipe, the heat dissipation profile remains unchanged with increasing $Ra$, which indicates conductive behavior dominating the region below the pipe. The heat distribution above the cable varies as the Rayleigh number increases due to buoyancy forces. Lower $Ra$ conditions show a lower rate of dissipation (concave profile), whereas soils experiencing a higher Rayleigh number show an accelerated heat distribution (convex profile). The heat profile above the cables transitions from concave to convex between a Rayleigh number of 2 and 7, below the established critical Ra of 12~\cite{nield1968onset}.

We closely examine the heat distribution above and below the pipeline to determine different modes of heat transfer.~\Cref{fig:heat-pipe} shows the evolution of temperature in the pipe and above the cables for different Rayleigh numbers. For $Ra < 1$, the average pipe temperature remains unchanged at 73.5\% of the cable temperature, and the soil temperature above the cable stays constant at 28\% of the cable temperature. As the Rayleigh number increases from 1 to 100, the temperature in the pipe drastically drops to 13\%, while the soil temperature rises to 95\% of cable temperature. The heating of pipe using cables is nonexistent for $Ra > 100$, where convection transfers almost all of the cable heat to the soil above. Hence, the transition from conduction to convection happens below a Rayleigh number of 1.0, much earlier than previously thought.

>>>>

As the permeability decreases from $k =$ \SI{4.9E-10} to \SI{1.6E-11}{\meter\squared}, the heat distributes more uniformly around the cable, heating the pipe and the fluid; but still showing a preferential upward heat distribution reaching the ground surface typical in a convection-dominated behavior. The pipe temperature increases showing symmetric heat distribution around the cable as the permeability decreases from \SI{1E-12} to \SI{1E-15}{\meter\squared}. In these low permeability conditions, conduction dominates, and the pipeline is heated efficiently, with minimal heat loss to the surrounding soil. At very low permeabilities (in silty and clayey soil with $k < \SI{1E-12}{\meter\squared}$ and $d_m < \SI{70}{\micro\meter}$), the narrow and sparsely interconnected pore spaces limit the fluid flow velocity. Hence, the time scale of convection is much larger than that of conduction, i.e., a smaller Rayleigh number indicating conduction-driven heat transfer. Conductive heat transfer causes an approximately isotropic distribution of heat around the source.

@@

\begin{figure}
 \centering
 \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{figs/ra/955.png}
    \caption{$Ra$ = 955, $k$ = \SI{1.91E-9}{\meter\squared}}
 \end{subfigure}
 \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{figs/ra/269.png}
    \caption{$Ra$ = 269.5, $k$ = \SI{5.64E-10}{\meter\squared}}
 \end{subfigure}\\
 \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{figs/ra/112.png}
    \caption{$Ra$ = 112, $k$ = \SI{2.47E-10}{\meter\squared}}
 \end{subfigure}
 \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{figs/ra/21.png}
    \caption{$Ra$ = 21.95, $k$ = \SI{4.89E-11}{\meter\squared}}
 \end{subfigure}\\ 
 \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{figs/ra/7.png}
    \caption{$Ra$ = 7.29, $k$ = \SI{1.67E-11}{\meter\squared}}
 \end{subfigure}
 \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{figs/ra/2.png}
    \caption{$Ra$ = 2.225, $k$ = \SI{4.84E-12}{\meter\squared}}
 \end{subfigure}\\  
\end{figure}%
\begin{figure}[ht]\ContinuedFloat 
 \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{figs/ra/0p45.png}
    \caption{$Ra$ = \SI{4.58E-1}, $k$ = \SI{1.05E-12}{\meter\squared}}
 \end{subfigure}
 \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{figs/ra/0p049.png}
    \caption{$Ra$ = \SI{4.9E-2}, $k$ = \SI{1.11E-13}{\meter\squared}}
 \end{subfigure}\\    
 \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{figs/ra/0p0049.png}
    \caption{$Ra$ = \SI{4.4E-4}, $k$ = \SI{1.05E-14}{\meter\squared}}
 \end{subfigure}
 \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{figs/ra/0p00049.png}
    \caption{$Ra$ = \SI{4.4E-4}, $k$ = \SI{1.18E-15}{\meter\squared}}
 \end{subfigure}\\ 
 \caption{Temperature distribution in different soil permeability conditions after 200 hours. The cables are heated to an above ambient temperature of \SI{70}{\celsius} and the burial depth is \SI{0.5}{\meter}.}
 \label{fig:ra-evol}
\end{figure}

The heat transfer mechanism changes from conduction to convection-dominated as the permeability of soil increases. A critical Rayleigh-Darcy number defines this transition point from conduction to convection.~\citeN{nield1968onset} defines this transition at a critical $Ra$ of 12 for a rectangular domain with a constant heat source with Free-Free boundary condition. At low permeabilities, where $Ra < 1$, the temperature field is symmetric, indicating a conductive behavior (see~\cref{fig:ra-evol} g-j). The asymmetric convective behavior occurs for $Ra > 2.22$ ($k > \SI{4.84E-12}{\meter\squared}$). The observation of convective response for $Ra = 2.22$, below the critical $Ra$ of 12, indicates the transition occurs at lower $Ra_c$ than originally thought.

\Cref{fig:ra-vline} shows the temperature distribution with depth along the A-A' axis for different Rayleigh-Darcy numbers. The fluid temperature in the pipe reaches a maximum of 73.5\% of the cable temperature ($T_{cable}$) for the lowest $Ra = $\SI{4.5E-4} ($k =$ \SI{1.18E-15}{\meter\squared}). The fluid temperature inside the pipe remains almost unchanged with increasing Rayleigh-Darcy number until a $Ra = 0.46$ ($k >$ \SI{1E-12}{\meter\squared}), above which the fluid temperature drops drastically. For large Rayleigh-Darcy numbers ($Ra > 100$), the fluid in the pipe barely reaches 5\% of the cable temperature. Below the pipe, the heat dissipation profile remains unchanged with increasing $Ra$, which indicates conductive behavior dominating the region below the pipe. The heat distribution above the cable varies as the Rayleigh-Darcy number increases due to buoyancy forces. Lower $Ra$ conditions show a lower rate of dissipation (concave profile), whereas soils experiencing a higher Rayleigh-Darcy number show an accelerated heat distribution (convex profile). The heat profile above the cables transitions from concave to convex between a Rayleigh number of 2 and 7, below the established critical Rayleigh-Darcy ($Ra_c$) of 12~\cite{nield1968onset}.

We closely examine the heat distribution above and below the pipeline to determine different modes of heat transfer.~\Cref{fig:heat-pipe} shows the evolution of normalized average temperatures of the fluid in the pipe, the soil below the pipe, and soil above the cables for different Rayleigh numbers. For $Ra < 1$, the average pipe temperature remains unchanged at 73.5\% of the cable temperature, and the soil temperature above the cable stays constant at 28\% of the cable temperature. As the Rayleigh number increases from 1 to 100, the temperature in the pipe drastically drops to 13\%, while the soil temperature rises to 95\% of cable temperature. The heating of pipe using cables is nonexistent for $Ra > 100$, where convection transfers almost all of the cable heat to the soil above. Hence, the transition from conduction to convection happens below a Rayleigh number of 1.0, much earlier than previously thought.
